---
title: "Exploring Data from the Lahman Package"
author: "Jim Albert albert@bgsu.edu"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning = FALSE,
                      message = FALSE)
```

#### R Markdown

In RStudio, a R Markdown file is a convenient way to combine text and R work.

#### Where is the R Code?

All of the R code for this workshop (as R Markdown files) will be available at

bit.ly/Rbaseball

#### Questions?

You are welcome to ask question in the Chat window. I'll answer some of these at the end of this 50-minute presentation.

#### The Data?

All of the data is contained as data frames within the Lahman package.

#### Packages

For this activity, I'll load three packages. (I am assuming that you already have them installed in R.)

```{r}
library(dplyr)
library(ggplot2)
library(Lahman)
```

#### Introduction to data wrangling

-   Start with a data frame - a rectangular array of data where rows correspond to observations and columns correspond to variables.

-   These dplyr verbs operate on a data frame. Use the piping operation ( %\>% ) to pass along the data frame between operations.

#### Common dplyr verbs:

-   filter() - choose rows
-   select() - choose columns
-   mutate() - create new variables
-   arrange() - sort by one or more variables
-   group_by() - divide data frame into groups
-   summarize() - compute a particular summary

#### Home Run Rate

Let's focus on home run rates, specifically the rate of home runs (HR) on balls in play (BIP): $$
Rate = \frac{HR}{BIP} = \frac{HR}{AB - SO}
$$

#### Best Single Season HR Rates?

In the last 60 seasons (1960 - 2019), let's find the top ten hitters with respect to single season home run rate. (At least 400 AB.)

#### Plan

-   Start with Batting data frame.
-   Use filter() function to only consider seasons 1960 or later.
-   For each player for each season, want to find the sum of AB, SO, HR.
-   Use mutate() function to compute the home run rate.
-   Store results in data frame S.

```{r}
Batting %>% 
  filter(yearID >= 1960) %>% 
  group_by(playerID, yearID) %>% 
  summarize(AB = sum(AB),
            SO = sum(SO),
            HR = sum(HR),
            .groups = "drop") -> S
```

Look at a few rows of S.

```{r}
head(S)
```

-   I'll limit my search to at least 400 AB.
-   Compute HR rates for all players.

```{r}
S %>% 
  filter(AB >= 400) %>% 
  mutate(Rate = HR / (AB - SO)) -> S400
```

-   Arrange by HR rate.
-   Display the top 10 HR rates.

```{r}
S400 %>% 
  arrange(desc(Rate)) %>% 
  head(10) -> topten
topten
```

Let's improve this display:

-   Add player names.
-   Only show name, yearID, HR, and Rate
-   round Rate to three decimal points.

```{r}
inner_join(select(Master, playerID, nameFirst,
                  nameLast), topten,
           by = "playerID") %>% 
          arrange(desc(Rate)) -> topten
topten
```

```{r}
topten %>% 
  mutate(Rate = round(Rate, 3)) %>% 
  select(nameFirst, nameLast, yearID, Rate) ->
  topten
topten
```

#### Graph Results

```{r}
ggplot(topten, aes(yearID, Rate)) +
  geom_point()
```

```{r}
ggplot(topten, aes(yearID, Rate, label = nameLast)) +
  geom_label()
```

#### Season Home Run Rates

Let's explore the pattern of overall home run rates over the past 50 seasons.

```{r}
Teams %>% 
  filter(yearID >= 1960) %>% 
  group_by(yearID) %>% 
  summarize(HR = sum(HR),
            AB = sum(AB),
            SO = sum(SO),
            Rate = HR / (AB - SO),
            .groups = "drop") -> S3
head(S3)
```

```{r}
ggplot(S3, aes(yearID, Rate)) +
  geom_point() +
  geom_smooth(span = 0.2)
```

#### Ohio Colleges

What Ohio colleges are represented in the Lahman database?

```{r}
Schools %>% 
  filter(state == "OH") %>% 
  pull(schoolID) -> ohio_schools
```

Which colleges had the most seasons containing future MLB players?

```{r}
CollegePlaying %>% 
  filter(schoolID %in% ohio_schools) %>% 
  group_by(schoolID) %>% 
  summarize(N = n(), .groups = "drop") %>% 
  arrange(desc(N)) %>% 
  head()
```

Among all Ohio players, who had the most career hits in the MLB?

Collect ids of all Ohio players.

```{r}
CollegePlaying %>% 
  filter(schoolID %in% ohio_schools) %>% 
  pull(playerID) %>% unique() -> ohio_players
```

Collect the number of MLB career hits for all Ohio players. Sort from highest to lowest.

```{r}
Batting %>% 
  filter(playerID %in% ohio_players) %>% 
  group_by(playerID) %>% 
  summarize(H = sum(H), .groups = "drop") %>% 
  arrange(desc(H)) %>% head(5) -> top_hits
```

Add first and last names to the list.

```{r}
inner_join(select(Master, playerID, nameFirst,
                  nameLast), top_hits,
           by = "playerID") %>% 
          arrange(desc(H)) 
```

Ohio player with the most MLB pitcher wins?

```{r}
Pitching %>% 
  filter(playerID %in% ohio_players) %>% 
  group_by(playerID) %>% 
  summarize(W = sum(W), .groups = "drop") %>% 
  arrange(desc(W)) %>% head(5) -> top_wins
```

Add first and last names.

```{r}
inner_join(select(Master, playerID, nameFirst,
                  nameLast), top_wins,
           by = "playerID") %>% 
          arrange(desc(W)) 
```

What Ohio schools did these pitchers attend?

```{r}
CollegePlaying %>% 
  filter(playerID %in% top_wins$playerID) %>% 
  group_by(playerID) %>% 
  summarize(College = first(schoolID),
            .groups = "drop")
```

#### World Series

What teams have won the most World Series?

```{r}
SeriesPost %>% 
  filter(round == "WS") %>% 
  group_by(teamIDwinner) %>% 
  summarize(N = n()) %>% 
  arrange(desc(N)) %>% 
  head()
```

What seasons did the Yankees win the WS?

```{r}
SeriesPost %>% 
  filter(round == "WS", 
         teamIDwinner == "NYA") %>% 
  pull(yearID)
```

#### Some Questions to Explore

1.  Suppose you define the walk rate to be $$
    BB Rate = \frac{BB}{AB + BB}
    $$

In the last 60 seasons (1960-2019), find the ten players that had the highest walk rates (assuming AB + BB >= 400).

2.  When did these top walk rate players play? (Do you see any pattern?)

3.  The demise of triples. It seems that triples are really rare in current MLB. Define the in-play triple rate to be 

$$
    Rate = \frac{Triples}{AB - SO}
$$ 
Find the triple rate for all seasons from 1960 through 2019. Construct a graph plotting the triple rate against season and add a smoothing curve. Do you see a pattern?

4.  Both triples and stolen bases are considered speed plays. Construct a scatterplot of the triple rate against the number of stolen bases for all seasons since 1960. Do you see a relationship?
