---
title: "Exploring Statcast Data for Four Pitchers"
author: "Jim Albert albert@bgsu.edu"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

#### Introduction

- Illustrate the Statcast database which gives much information about every pitch thrown in a MLB game

- I have created a subset of the 2020 Statcast data, focusing on four starting pitchers

- Focus on variables that describe the pitch type, speed, break, location, and outcome (swing and miss) for each pitch

#### R Code?

- All of the R code for this workshop (as R Markdown files) will be available at

bit.ly/Rbaseball

#### Questions?

You are welcome to ask question in the Chat window.  I'll answer some of these at the end of this 50-minute presentation.

#### Read in four packages

```{r}
library(dplyr)
library(readr)
library(ggplot2)
library(janitor)
```

#### Source in a file to add a zone to a graph

```{r}
source("http://bit.ly/add_zone")
```

#### Data

I have collected Statcast data for all pitches for four pitchers for the 2020 season.

```{r}
d <- read_csv("http://bit.ly/4pitchers")
```

By the way, I added three new variables to the dataset:

- Pitcher - last name of the pitcher
- Swing - did the batter swing?  1 (yes), 0 (no)
- Miss - did the batter miss?  1 (yes), 0 (no)

#### Who Are the Pitchers?

What pitchers did I include?

```{r}
d %>% 
  tabyl(Pitcher)
```

#### Types of pitches

A nice reference:  Terry McDermott _Off Speed:  Baseball, Pitching \& and the Art of Deception_


- (FF) Four-Seam Fastball: A straight fastball, almost always the fastest ball any pitcher can throw. Speeds range from low 80's to 100 mph.

- (FT) Two-Seam Fastball:  Fastball gripped with the seams rather than across them.  Sinks and often moves left to right from a right-handed pitcher.

- (CB and KC) Curveball:  Most basic breaking pitch.  Pitcher imparts spin to the ball, causing it to break both horizontally right-to-left and down (for a right-handed pitcher).

- (SL) Slider:  Faster breaking ball than curveball.  Breaks more horizontally and less vertically then a normal curveball.  

- (FC) Cut Fastball:  Cutter lies halfway along the continuum between a slider and a fastball.  Relatively little break, but does this late.

- (CH) Changup:  Thrown like a fastball but leaves the hand as much as 10 mph slower.  Often imparts a slight right-to-left break from a right-handed pitcher.

- (SI) Sinker:  Similar to a two-seam fastball.

- (FS) Splitter:  Similar to a sinker but with a much sharper downward break.  Slower than a fastball.

#### What pitches did our pitchers throw?

```{r}
d %>% 
  tabyl(pitch_name)
```

Breakdown pitch types by pitcher:

```{r}
d %>% 
  tabyl(pitch_name, Pitcher)
```

#### Four-Seam Fastball Speeds

```{r}
ggplot(filter(d, pitch_type == "FF"),
       aes(release_speed)) +
  geom_histogram(color = "white",
                 fill = "blue")
```

How do these fastball speeds vary by pitcher?

```{r}
ggplot(filter(d, pitch_type == "FF"),
       aes(release_speed)) +
  geom_histogram(color = "white",
                 fill = "blue") +
  facet_wrap(~ Pitcher, ncol = 1)
```

Average four-seam fastball speed of each pitcher?

```{r}
d %>% 
  filter(pitch_type == "FF") %>% 
  group_by(Pitcher) %>% 
  summarize(Count = n(),
            Mean = mean(release_speed))
```

#### Movement

Let's focus on Kershaw.  Compare the movement of each type of pitch.

```{r}
ck <- filter(d, Pitcher == "Kershaw")
```

```{r}
ggplot(ck, 
       aes(pfx_x, pfx_z, 
           color = pitch_type)) +
  geom_point()
```

#### Location of Pitches?

```{r}
ggplot(ck, 
       aes(plate_x, plate_z, 
           color = pitch_type)) +
  geom_point()
```

```{r}
ggplot(ck, 
       aes(plate_x, plate_z, 
           color = pitch_type)) +
  geom_point() +
  add_zone()
```

```{r}
ggplot(ck, 
       aes(plate_x, plate_z)) +
  geom_point() +
  add_zone() +
  facet_wrap(~ pitch_type, ncol = 2)
```

#### Swing and Miss

Focus on 

- Kershaw's curve balls
- Batter swings at pitch (Swing = 1)
- Record if there is a Miss (Miss = 1) or 
Contact (Miss = 0)

```{r}
ggplot(filter(ck, pitch_type == "CU",
              Swing == 1),
       aes(plate_x, plate_z, color = Miss)) +
  geom_point() + add_zone()
```

What fraction of swings on curveballs were missed?

```{r}
ck %>% 
  filter(pitch_type == "CU",
         Swing == 1) %>% 
  tabyl(Miss)
```

How does that compare with other pitchers?

Compare the curveballs of Kershaw and Wheeler.

```{r}
d %>% 
  filter(pitch_type == "CU",
         Swing == 1) %>% 
  tabyl(Pitcher, Miss) %>% 
  adorn_percentages("row")
```

Compare the Knuckle Curves of Bauer and Cole.

```{r}
d %>% 
  filter(pitch_type == "KC",
         Swing == 1) %>% 
  tabyl(Pitcher, Miss) 
```

Maybe Cole is better because of its movement?

```{r}
d %>% 
  filter(pitch_type == "KC") %>% 
  ggplot(aes(pfx_x, pfx_z,
             color = Pitcher)) +
  geom_point()
```


Maybe Cole is better because of its location?

```{r}
d %>% 
  filter(pitch_type == "KC") %>%
  ggplot(aes(plate_x, plate_z)) +
  geom_point() + 
  add_zone() +
  facet_wrap(~ Pitcher)
```

Maybe Cole is better since he has a better spin rate?

```{r}
d %>% 
  filter(pitch_type == "KC") %>% 
  ggplot(aes(Pitcher, release_spin_rate)) +
  geom_jitter() 
```

Relationship between spin and movement?

Focus on Cole's knuckle curveball.

```{r}
ggplot(filter(d, pitch_type == "KC",
              Pitcher == "Cole"),
  aes(release_spin_rate, pfx_z)) +
  geom_point() +
  geom_smooth()
```

```{r}
ggplot(filter(d, pitch_type == "KC",
              Pitcher == "Cole"),
  aes(release_spin_rate, pfx_x)) +
  geom_point() +
  geom_smooth()
```












